Using a Deep Autoencoder Model to Characterize Air Pollution

### 2021-09-08 paper-draft/paper.tex
Updated code

\title{Using a Deep Autoencoder Model to Characterize Air Pollution}
\author{Nicolas Shannon$^{1}$, Bishnu Timalsena$^{2}$, Raul Chavez Jr.$^{3}$, \\Vandana Nunna Lakshmi$^{2}$}
\date{}
% To cite, use \parencite{} or \textcite{}

@@

% Acknowledgements
\begin{centering}
\maketitle
\vspace{-0.50cm}
\scriptsize
{$^{1}$Department of Computer Science, Loyola University Chicago, Chicago, IL, US}\\
{$^{2}$Department of Computer Science, University North Texas, Denton, TX, US}\\
{$^{3}$Department of Electrical and Computer Engineering, University of Texas 
    - Rio Grande Valley, Brownsville, TX, US \note{$\leftarrow$ fix formatting}}
\end{centering}

@@

\subsection{A Deep Learning Approach}
In this paper we compare the autoencoder model with principle component analysis (PCA), a much
older technique for dimensional reduction. Using a deep autoencoder has several advantages.
First, deep autoencoders are better at compressing highly complex data into few dimensions.
Autoencoders with a nonlinear encoder and decoder function can learn more powerful nonlinear
generalizations than PCA. \parencite{Goodfellow16}.
\par \note{Autoencoder grapic from poster here}

\section{Related Work}
The idea of generating embeddings for a location is a highly potential expanse for research.
The two major forces behind its competency are, the fact that large volumes of data is being
effectively compressed and that these embeddings, when plugged into any model could effectively
represent any given location.  A lot of research is being done on location embeddings in recent
times. For instance, location embeddings are being generated based on human-mobility data 
(Crivellari, A.; 2019) to understand the behavioral proximity between different locations 
irrespective of their spatial distances. The sequence of locations that are frequently
travelled were fed to a Skip-gram Word2vec model to generate the location embeddings. Another 
approach that was implemented to leverage location embeddings is, GPS2Vec ( Yifang Yin, 2019).
This application consists of a neural network that would generate embeddings for a location, 
based on the semantic contexts that are sourced from the user content published on multiple 
digital platforms. With the integration of these embeddings, a successful geo-tagged image
classification task is demonstrated. In another effort (Carl Yang, 2019), the location 
embeddings derived from user content such as addresses, phone numbers, images, place names and 
all other details from multiple individual online posts, are compressed to form low dimensional
embeddings to solve the place duplication issue. Deduplication of locations, is a paramount 
task for any digital platform since they provide a place graph to its users, which is the 
result of merging location data from multiple sources. This merge leads to duplicity since 
different sources could hold different names for the same place. By applying K-NN search, 
pair-wise duplication prediction on these generated location embeddings, the application can
identify the duplicates and eliminate them to result in an efficient place graph.

@@

We used two different methods to reduce the dimensionality of our data - Principle Component
Analysis and Undercomplete Autoencoders. A simple linear regression was performed on 
the encoded data that was generated by both the PCA and AE models. Additionally, we performed
a linear regression on the unencoded values for comparison. Each model's goodness of fit
was evaluated using the percentage of explained variance for the three linear regressions.

@@

\subsection{Data Collection and Preprocessing}
The dataset was created using OpenWeather's Air Pollution API. The dataset includes data for 
eight polluting gases and particulates: CO, NO, NO$_2$, O$_3$, SO$_2$, NH$_3$, PM$_{2.5}$, 
and PM$_{10}$. The dataset also includes the hourly timestamp, the latitude and longitude (part of the input),
and the air quality index, the latter of which we did not end up using.\par
We initially looked at data for the eight pollutants over a six-and-a-half month time 
frame: January 27th, 2020 through July 6th, 2021. However, we found that we achieved
better results with just four months of of data for our input features. Our final 
dataset included just over eighteen thousand cities in the United States and other U.S. 
territories. Additionally, we found it easier to focus on a single gas when developing 
the models, so we chose to concentrate on CO during the engineering process. Given
enough time and computational power, the techniques performed on the CO model would be 
easily extensible and transferable to the other seven pollutants.

% dataset characteristics table
\begin{table}[h!]
    \caption{Dataset Characteristics}
    \label{tab:table1}
    \vspace{0.1cm}
    \begin{tabular}{p{4cm}p{7cm}}
        \hline
        \multicolumn{2}{c}{OpenWeather Pollution API} \\
        \hline
        Time period  & 6.5 months    \\
        Time frequency & Daily  \\
        $\#$ of cities & 18,526 \\
        $\#$ of features & 120 (originally 190) \\
        $\#$ of component gases & 8 \\
        Dependent variable & Single day of time series (3/27/21) \\
        \hline
    \end{tabular}
\end{table}

OpenWeather's Pollution API ordinarily returns hourly values, but we opted to use the daily
average pollution levels as our feature set. We resolved to feature engineer the daily 
averages and to reduce the time scale to four months in order to keep the data 
proportional to the number of features and thus avoid the curse of dimensionality. 
\parencite{Trunk79}. \note{Idk about this} \par
We separated the dependent from the input before unit normalizing the input. 
The normalized data was then encoded by each model.

\subsection{Autoencoder and PCA Models}
The autoencoder model used consists of an encoded and decoded layer. We performed
a grid search for the CO model to determine the optimal hyperparameters. The following 
hyperparameters were tested:
\begin{itemize}
    \item \textit{Learning Rate:} 0.0001, 0.001, 0.01, 0.1
    \item \textit{Batch Size:} 32, 64, 128, 256
    \item \textit{Epochs:} 10, 50, 75, 100
\end{itemize}
We used the parameters that appeared most frequently over the one hundred twenty dimensions and 
applied them over the entire range so as not to overfit.\note{$\leftarrow$ not sure if this is correct}

\begin{table}[h!]
    \caption{Dataset Characteristics}
    \label{tab:table1}
    \vspace{0.1cm}
    \begin{tabular}{p{4cm}p{7cm}}
        \hline
        \multicolumn{2}{c}{Optimal Hyperparameters for Carbon Monoxide (CO)} \\
        \hline
        Activation function & Leaky ReLU \\
        Loss function & Mean Squared Error  \\
        Optimizer & Adam \\
        Learning rate & 0.001 \\
        Batch size & 64 \\
        Epochs & 150 \\ 
        \hline
    \end{tabular}
\end{table}

The PCA model was trained with k-fold cross-validation using five folds. The dataset was
split into a training, validation, and hold-out test set.

>>>>

\title{Using a Deep Autoencoder Model to Characterize Air Pollution}
\author{Nicolas Shannon$^{1}$, Vandana Nunna Lakshmi$^{2}$}
\date{}
% To cite, use \parencite{} or \textcite{}

@@

% Acknowledgements
\begin{centering}
\maketitle
\vspace{-0.50cm}
\scriptsize
{$^{1}$Department of Computer Science, Loyola University Chicago, Chicago, IL, US}\\
{$^{2}$Department of Computer Science, University North Texas, Denton, TX, US}\\
%{$^{3}$Department of Electrical and Computer Engineering, University of Texas 
%    - Rio Grande Valley, Brownsville, TX, US \note{$\leftarrow$ fix formatting}}
\end{centering}

@@

\subsection{A Deep Learning Approach}
In this paper we employ autoencoder and  principle component analysis (PCA) models to 
feed the pollutant data and perform data compression. The resulting embeddings are 
compared in terms of effeciency and data loss. The features are encoded at different
scales of dimensions to check the fluctuation in explainable variance when predicting a 
given target. Given that each feature in the dataset represents the daily average of a 
specific pollutant in that location and that pollution levels do not vary drastrically 
on a day-to-day basis, it is evident that there is a strong correlation among features
\par While a healthy correlation among features is encouraged, abnormally strong values
relfect redundancy and end up being a burden on the model.
\parencite{featureredundancy}. First, deep autoencoders are better at compressing highly 
complex data into fewer dimensions. Autoencoders with a nonlinear encoder and decoder 
function can learn more powerful nonlinear generalizations than PCA. 
\parencite{Goodfellow16}.

\section{Related Work}
The idea of generating embeddings for a location is a highly potential expanse for
research. The two major forces behind its competency are, the fact that large volumes of
data is being effectively compressed and that these embeddings, when plugged into any
model could effectively represent any given location.  A lot of research is being done 
on location embeddings in recent times. For instance, location embeddings are being 
generated based on human-mobility data \parencite{GeoEmbeddings} to understand the 
behavioral proximity between different locations irrespective of their spatial distances.
The sequence of locations that are frequently travelled were fed to a Skip-gram Word2vec 
model to generate the location embeddings. Another approach that was implemented to 
leverage location embeddings is, GPS2Vec  \parencite{GPS2Vec}.  This application consists
of a neural network that would generate embeddings for a location, based on the semantic 
contexts that are sourced from the user content published on multiple digital platforms.
With the integration of these embeddings, a successful geo-tagged image classification 
task is demonstrated. In another effort  \parencite{PlaceDeduplication}, the location 
embeddings derived from user content such as addresses, phone numbers, images, place
names and all other details from multiple individual online posts, are compressed to 
form low dimensional embeddings to solve the place duplication issue. Deduplication of 
locations, is a paramount task for any digital platform since they provide a place graph
to its users, which is the result of merging location data from multiple sources. This 
merge leads to duplicity since different sources could hold different names for the same 
place. By applying K-NN search, pair-wise duplication prediction on these generated 
location embeddings, the application can identify the duplicates and eliminate them to 
result in an efficient place graph.\par
In 2018, a new approach, “DeepMove” is proposed \parencite{DeepMove}. This 
implementation employs Trip model and OD model to learn latent representations of 
places.  The approach is successfully applied to New York city yellow taxi dataset – 
“includes pick-up/drop-off times, the number of passengers, pick-up/drop-off locations 
represented by longitude and latitude, trip distance, rate types, payment type, and
itemized fares” \parencite{DeepMove}. Similarities among locations in New York city are 
derived effectively with “DeepMove”. To resolve the dissonance that arises from the 
fusion of geographic and semantic features of a location, an unsupervised learning 
implementation had been demonstrated that would derive location embeddings from human 
trajectories \parencite{DisentangledLocationEmbeddings}. With the ever- growing 
footprint of digital and social media platforms in everyday life of an individual, a
lot of research is happening to generate embeddings based on the content posted in 
different locations. LeGo-CM is one such effort that would generate embeddings of 
geotagged tweets \parencite{GeotaggedTweets}. These resultant embeddings could be 
applied to understand the kind of content each location reflects.Venue2Vec is another 
idea that is implemented to enhance the location-aware services in various platforms 
\parencite{Venue2Vec}. This model is trained on semantic data, temporal-spatial context 
and sequential relations among locations to generate embeddings. This makes sure that 
locations with similar data would fall close, when plotted in a multi-dimensional space,
based on the input features. Hence, the model is effective in predicting the next and  
future check-in location. Community Flow prediction is another area of interest where
embeddings are a major problem solver. Geo-contextual Multitask Embedding Learner (GMEL)
, is a model that uses “Graph Attention Network” to encode geographical contexts and 
spatial correlations to generate location embeddings \parencite{CommutingFlowPrediction}.
These embeddings are further used to predict the community fow that serves as a vital
factor in policy establishment and metro planning.

@@

We used two different methods to reduce the dimensionality of our data - Principle Component
Analysis and undercomplete autoencoders. A simple linear regression was performed on 
the encoded data that was generated by both the PCA and AE models. Additionally, we performed
a linear regression on the unencoded values for comparison. Each model's goodness of fit
was evaluated using the percentage of explained variance for the three linear regressions.

@@

\subsection{Data Collection and Preprocessing}
The dataset was created using OpenWeather's Air Pollution API. The dataset includes data for 
eight polluting gases and particulates: CO, NO, NO$_2$, O$_3$, SO$_2$, NH$_3$, PM$_{2.5}$, 
and PM$_{10}$. The dataset also includes the hourly timestamp, the latitude and longitude (part of the input),
and the air quality index, the latter of which we did not end up using.\par
We looked at data for the eight pollutants over a six-and-a-half month time 
frame: January 27th, 2020 through June 6th, 2021. Our final 
dataset included just over eighteen thousand cities in the United States and other U.S. 
territories. We found it easier to focus on a single gas when developing 
the models, so we chose to concentrate on CO during the engineering process. Given
enough time and computational power, the techniques performed on the CO model would be 
easily extensible and transferable for the other seven pollutants.

% dataset characteristics table
\begin{table}[h!]
    \caption{Dataset Characteristics}
    \label{tab:table1}
    \vspace{0.1cm}
    \begin{tabular}{p{4cm}p{7cm}}
        \hline
        \multicolumn{2}{c}{OpenWeather Pollution API} \\
        \hline
        Sampling period  & 11/27/20 to 06/05/21\\
        Time interval & Daily  \\
        $\#$ of cities & 18,526 \\
        $\#$ of features & 191 \\
        $\#$ of component gases & 8 \\
        Dependent variable & Single day of time series (06/05/21) \\
        \hline
    \end{tabular}
\end{table}

OpenWeather's Pollution API ordinarily returns hourly values, but we opted to use the daily
average pollution levels as our feature set in order to make the number of starting dimensons
more manageable. We resolved to feature engineer the daily 
averages and proportional to the number of features and thus avoid the curse of dimensionality. 
\parencite{Trunk79}. \note{Idk about this} \par
We separated the dependent from the input before unit normalizing the input. 
The normalized data was then encoded by each model.

\subsection{Undercomplete Autoencoder Model}
The autoencoder model used consists of an encoded and decoded layer. We performed
a grid search for the CO model to determine the optimal hyperparameters. The following 
hyperparameters were tested:
\begin{itemize}
    \item \textit{Learning Rate:} 0.0001, 0.001, 0.01, 0.1
    \item \textit{Batch Size:} 32, 64, 128, 256
    \item \textit{Epochs:} 10, 50, 75, 100
\end{itemize}
K-folds cross validation was used to validate the grid search using five folds. We selected the
highest coefficient of determination that appeared in each of the five folds for every given
dimension. The most frequently occuring combination of hyperparameters was then used to create
the finalized embedding.

\begin{table}[h!]
    \caption{Dataset Characteristics}
    \label{tab:table1}
    \vspace{0.1cm}
    \begin{tabular}{p{4cm}p{7cm}}
        \hline
        \multicolumn{2}{c}{Optimal Hyperparameters for Carbon Monoxide (CO)} \\
        \hline
        Activation function & tanh\\
        Loss function & Mean Squared Error  \\
        Optimizer & Adam \\
        Learning rate & 0.001 \\
        Batch size & 64 \\
        Epochs & 150 \\ 
        \hline
    \end{tabular}
\end{table}

\subsection{Principle Component Analysis}
The PCA model was trained with k-fold cross-validation using five folds. The dataset was
split into a training, validation, and hold-out test set.

\subsection{Model Validation}
K-folds cross validation was performed on the linear regression test using five folds. The folds
with the best and worst variance scores were extracted so as to make an equitable evaluation of
the model.
